Efficient pedigree recording for fast population genetics simulation

### 2018-01-14 forwards_paper.tex
factors of 2

%% Progressive memory reduction as a function of T
\citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
by retaining only 8 generations of history.
This begs the question: how fast does the information in the pedigree become irrelevant?
Suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
One way to answer this question is to ask:
How many of the resulting $4NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $2N-2$
plus four times the sum of the lengths of all marginal trees,
because new edges appear with most recombinations that occur on any marginal tree.
Therefore, we need to compute the mean total length of all branches in a coalescent tree
up until time $T$ (or the common ancestor, whichever comes first).
Since $k$ lineages coalesce at rate $\binom{k}{2}/N$,
the expected length of time for which a marginal tree has $k$ branches is
$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
Since there are $k$ branches for this time,
this period is expected to contribute $2N/(k-1)$ generations to the total tree length.
By summing over $n < k \le N$,
the $N$ tips of a tree are expected to descend from only $n$ lineages
around $2N(1/(n-1) - 1/(N-1))$ generations ago.
Inverting this relationship between time and number of roots
implies that a marginal tree cut $T$ units of time ago
is expected to have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align}
These calculations hold up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.

\begin{figure}
    \begin{center}
        \includegraphics{sims/simplify-results}
    \end{center}
    \caption{
        Time and space complexity of simplify.
        \textbf{(A)}
        Number of edges in the simplified tree sequence
        for 10 replicate Wright--Fisher simulations with $N=50$ as a function
        of number of generations.
        Each line is one simulation, the heavy blue line gives the average,
        and the dashed line is the upper bound of equation \eqref{eqn:edge_bound}.
        \textbf{(B)}
        Time required to simplify the first $k$ edges of a large (4.2GiB)
        unsimplified tree sequence produced by a forwards-time simulation plotted
        against $k$. The time scales linearly with the number of input edges.
        \textbf{(C)}
        Time required to simplify the tree sequence resulting from a coalescent
        simulation of 500,000 samples of a 200 megabase human chromosome
        to a random subsample of $n$ samples, plotted against $n$
        (note the log scale; the time scales logarithmically with $n$).
        \label{fig:simplify_complexity}
    }
\end{figure}

%% Above was edges; same holds for memory usage savings due to mutations
If we were to add mutations in forwards time
under the infinite-sites model with total mutation rate per generation $\mu$ (i.e., per meiosis),
there would also be around $\mu 2NT$ mutations, and the same number of sites,
simply because there was a total of $2NT$ meioses.
Since mutations fall as recombinations do on the marginal trees,
the number of mutations is proportional to the total length of all the marginal trees,
and so by the same argument as for the number of edges,
adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations.
With $N=10^4$ and $T=20N$ as above,
this implies that adding neutral mutations afterwards reduces the number of mutations that must be generated
by a factor of 10,000.
This could result in substantial time savings,
even without considering the
computational burden of propagating mutations forwards across generations.

>>>>

%% Progressive memory reduction as a function of T
\citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
by retaining only 8 generations of history.
This begs the question: how fast does the information in the pedigree become irrelevant?
Suppose instead that we have run a simulation of $N$ individuals for $T$ generations,
begun with no prior history.
One way to answer this question is to ask:
How many of the resulting $2NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $N-2$
plus four times the sum of the lengths of all marginal trees,
because new edges appear with most recombinations that occur on any marginal tree.
Therefore, we need to compute the mean total length of all branches in a coalescent tree
up until time $T$ (or the common ancestor, whichever comes first).
Since $k$ lineages coalesce at rate $\binom{k}{2}/N$,
the expected length of time for which a marginal tree has $k$ branches is
$N/(k(k-1)) = N(1/(k-1) - 1/k)$ generations.
Since there are $k$ branches for this time,
this period is expected to contribute $N/(k-1)$ generations to the total tree length.
By summing over $n < k \le N$,
the $N$ tips of a tree are expected to descend from only $n$ lineages
around $N(1/(n-1) - 1/(N-1))$ generations ago.
Inverting this relationship between time and number of roots
implies that a marginal tree cut $T$ units of time ago
is expected to have around $n(T)$ roots, where $n(T) \approx 1 + N/(T+2)$.
The total tree length over this time is
$N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     N \log\left( \frac{N/2}{1 + N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    N \left( 1 + 4 \log\left( \frac{NT}{2(T + N)} \right)\right) .
\end{align}
These calculations hold up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.

\begin{figure}
    \begin{center}
        \includegraphics{sims/simplify-results}
    \end{center}
    \caption{
        Time and space complexity of simplify.
        \textbf{(A)}
        Number of edges in the simplified tree sequence
        for 10 replicate Wright--Fisher simulations with $N=100$ as a function
        of number of generations.
        Each line is one simulation, the heavy blue line gives the average,
        and the dashed line is the upper bound of equation \eqref{eqn:edge_bound}.
        \textbf{(B)}
        Time required to simplify the first $k$ edges of a large (4.2GiB)
        unsimplified tree sequence produced by a forwards-time simulation plotted
        against $k$. The time scales linearly with the number of input edges.
        \textbf{(C)}
        Time required to simplify the tree sequence resulting from a coalescent
        simulation of 500,000 samples of a 200 megabase human chromosome
        to a random subsample of $n$ samples, plotted against $n$
        (note the log scale; the time scales logarithmically with $n$).
        \label{fig:simplify_complexity}
    }
\end{figure}

%% Above was edges; same holds for memory usage savings due to mutations
If we were to add mutations in forwards time
under the infinite-sites model with total mutation rate per generation $\mu$ (i.e., per meiosis),
there would also be around $\mu NT$ mutations, and the same number of sites,
simply because there was a total of $NT$ meioses.
Since mutations fall as recombinations do on the marginal trees,
the number of mutations is proportional to the total length of all the marginal trees,
and so by the same argument as for the number of edges,
adding them after simplification results in only around $\mu 2 N \log(N)$ mutations.
With $N=2 \times 10^4$ and $T=10N$ as above,
this implies that adding neutral mutations afterwards reduces the number of mutations that must be generated
by a factor of 10,000.
This could result in substantial time savings,
even without considering the
computational burden of propagating mutations forwards across generations.
